Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-03-20 paper.qmd
Updated Figure 9

@@

gen_tuning_lg_plots()
```

>>>>

```{r}
#| label: model-two-curvy-umap

umap_two_curvy <- read_rds(file = "data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_umap_n-neigbors_15_min-dist_0.1.rds")

num_bins_x_two_curvy <- 42

algo_obj_two_curvy3 <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_two_curvy, 
  nldr_data = umap_two_curvy, 
  num_x_bins = num_bins_x_two_curvy,
  col_start = "x")

umap_two_curvy_scaled <- algo_obj_two_curvy3$nldr_scaled
distance_two_curvy <- algo_obj_two_curvy3$distance_df
tr_from_to_df_two_curvy <- algo_obj_two_curvy3$tr_from_to_df
benchmark_two_curvy <- algo_obj_two_curvy3$benchmark
df_bin_centroids_two_curvy <- algo_obj_two_curvy3$df_bin_centroids
df_bin_two_curvy <- algo_obj_two_curvy3$df_bin
hex_grid_with_counts <- algo_obj_two_curvy3$hex_grid_with_counts
a1 <- algo_obj_two_curvy3$a1

trimesh_two_curvy_umap <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000",
               alpha = 0.5,
               size = 0.5) +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  xlim(c(0, 1)) +
  ylim(c(0, 1)) +
  interior_annotation("a1", sc_ltr_pos, cex = 2)

```{r}

####With diff benchmark values

## Benchmark 2

benchmark_two_curvy <- 7 * a1

distance_df_small_edges_two_curvy3 <- distance_two_curvy |>
  filter(distance < benchmark_two_curvy)

tr_from_to_df_two_curvy4 <- inner_join(
  tr_from_to_df_two_curvy, distance_df_small_edges_two_curvy3,
  by = c("from", "to"))

trimesh_two_curvy_removed2_umap <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy4,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000",
               alpha = 0.5,
               size = 0.5)+
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  xlim(c(0, 1)) +
  ylim(c(0, 1)) +
  interior_annotation("b1", sc_ltr_pos, cex = 2)

benchmark_two_curvy <- 4 * bin_width

distance_df_small_edges_two_curvy2 <- distance_two_curvy |>
  filter(distance < benchmark_two_curvy) ## 0.231

tr_from_to_df_two_curvy3 <- inner_join(
  tr_from_to_df_two_curvy, distance_df_small_edges_two_curvy2,
  by = c("from", "to"))

trimesh_two_curvy_removed1_umap <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy3,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000",
               alpha = 0.5,
               size = 0.5)+
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  xlim(c(0, 1)) +
  ylim(c(0, 1)) +
  interior_annotation("c1", sc_ltr_pos, cex = 2)
```

```{r}
#| label: compare-2d-and-4d-dist

### Compute highD distance
dist_vec <- proxy::dist(x = df_bin_two_curvy[, -1], method = "Euclidean") |> as.vector()

from_vec <- c()
to_vec <- c()
num_obs <- 1:(NROW(df_bin_two_curvy) - 1)

for (obs in num_obs) {

  from_val <- rep(obs, (NROW(df_bin_two_curvy) - obs))
  if ((obs + 1) <= NROW(df_bin_two_curvy)) {
    to_val <- (obs + 1):NROW(df_bin_two_curvy)
  }
  from_vec <- append(from_vec, from_val)
  to_vec <- append(to_vec, to_val)

dist_highd <- tibble::tibble(from = from_vec, to = to_vec, dist_highd = dist_vec)

## To plot the distribution of distances

# Define bin width and starting point
start_point <- a1/2

# Create bins and calculate the mean value for each bin range
bins <- seq(start_point, max(distance_two_curvy$distance) + a1,
            by = a1)
bin_labels <- bins[-length(bins)] + a1 / 2  # mean of each bin range

distance_two_curvy_dist <- distance_two_curvy |>
  mutate(bin = cut(distance, breaks = bins, include.lowest = TRUE, labels = bin_labels))

distance_two_curvy_dist <- left_join(distance_two_curvy_dist, dist_highd, by = c("from", "to"))

text_df <- tibble::tibble(
  x = c(0.035, 0.15),
  y = c(10, 10),
  text = c("c", "b")
)

distance_points_umap <- ggplot(
  distance_two_curvy_dist,
  aes(x = distance,
      y = dist_highd)) +
  geom_point(alpha = 0.5) +
  geom_text(
    data=text_df,
    aes(x=x, y=y,
        label=text),
    colour="#bdbdbd",
    size = 10) +
  geom_vline(xintercept = 6.5 * bin_width,
             linetype="solid",
             color = "#bdbdbd",
             linewidth=1) +
  geom_vline(xintercept = 2 * bin_width,
             linetype="solid",
             color = "#bdbdbd",
             linewidth=1) +
  ylab(expression(d^{(4)})) +
  xlab(expression(d^{(2)})) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        panel.grid.major.x = element_blank(),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
```

```{r}
#| label: proj-tsne-model

# Apply the scaling
df_model_data_two_curvy <- bind_rows(data_two_curvy, df_bin_two_curvy)

proj_obj1 <- get_projection(projection = projection, 
                            proj_scale = 1, 
                            scaled_data = scaled_two_curvy_data, 
                            scaled_data_model = scaled_two_curvy_data_model, 
                            distance_df_small_edges = distance_df_small_edges_two_curvy, 
                            axis_param = list(limits = 0.6, 
                                              axis_scaled = 2,
                                              axis_pos_x = -0.5, 
                                              axis_pos_y = -0.5, 
                                              threshold = 0.04))

## Projection with Delauany triangulation

model_df1 <- dplyr::left_join(
  distance_two_curvy |> select(-distance),
  projected_model_df,
  by = c("from" = "ID"))

names(model_df1)[3:NCOL(model_df1)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_from")

model_df1 <- dplyr::left_join(model_df1, projected_model_df, by = c("to" = "ID"))
names(model_df1)[(2 + NCOL(projected_model_df)):NCOL(model_df1)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_to")

## With benchmark option 2

model_df2 <- dplyr::left_join(
  distance_df_small_edges_two_curvy3 |> select(-distance),
  projected_model_df,
  by = c("from" = "ID"))

names(model_df2)[3:NCOL(model_df2)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_from")

model_df2 <- dplyr::left_join(model_df2, projected_model_df, by = c("to" = "ID"))
names(model_df2)[(2 + NCOL(projected_model_df)):NCOL(model_df2)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_to")

two_curvy_proj_first_model1_umap <- plot_proj(
    projected_df = projected_df, 
    model_df = model_df, 
    axes = axes, 
    circle = circle, 
    point_param = c(0.5, 0.2, "#636363"), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-0.6, 0.6), 
    title = "c2", 
    cex = 2, 
    position = sc_ltr_pos,
    axis_text_size = 4, 
    is_category = FALSE) +
    theme(
        legend.position = "none"
    )

## Projection with Delauany triangulation

two_curvy_proj_model_delaunay_umap <- plot_proj(
    projected_df = projected_df, 
    model_df = model_df1, 
    axes = axes, 
    circle = circle, 
    point_param = c(0.5, 0.2, "#636363"), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-0.6, 0.6), 
    title = "a2", 
    cex = 2, 
    position = sc_ltr_pos,
    axis_text_size = 4, 
    is_category = FALSE) +
    theme(
        legend.position = "none"
    )

two_curvy_proj_model_benchmark2_umap <- plot_proj(
    projected_df = projected_df, 
    model_df = model_df2, 
    axes = axes, 
    circle = circle, 
    point_param = c(0.5, 0.2, "#636363"), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-0.6, 0.6), 
    title = "b2", 
    cex = 2, 
    position = sc_ltr_pos,
    axis_text_size = 4, 
    is_category = FALSE) +
    theme(
        legend.position = "none"
    )

@@

free(distance_points_umap) + 
    wrap_plots(trimesh_two_curvy_umap, trimesh_two_curvy_removed2_umap,
               trimesh_two_curvy_removed1_umap, two_curvy_proj_model_delaunay_umap,
               two_curvy_proj_model_benchmark2_umap, two_curvy_proj_first_model1_umap,
               ncol = 3) +
    plot_layout(heights = c(0.8, 1))
```
